Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-26 Texfiles/Draft1.tex
updates

\hypersetup{ colorlinks, citecolor=black, filecolor=black, linkcolor=black,
    urlcolor=black }
\setlength{\parindent}{0pt}
\usepackage[parfill]{parskip}
\date{July 2023}
\title{Retirement Consumption: Evidence from UK Pension Reform}
\author{INSERT NUMBER HERE}
\usepackage[letterpaper, margin=1.1in]{geometry}
\linespread{1.25}

\begin{document}

\newpage
\maketitle
\tableofcontents
\newpage

Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \cite{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts - this phenomenon has been termed the "annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature. [\textbf{Add
            something once I have a conclusion}].

@@

\cite{yaari_65} showed that under standard assumptions we would expect
individuals to annuitise all of their wealth at retirement to insure against the
risk of long life. Since then there has been much literature discussing possible
reasons that people do not annuitise. \cite{finkelstein_porteba_2002} and
\cite{finkelstein_porteba_2004} find evidence of adverse selection, thereby
making the 'money's worth' of annuities lower for the general population as
opposed to the population of annuitants. However, they also find that theory
would still predict annuitisation.

@@

\cite{lockwood_red_2012} builds on this and shows that a realistic bequest
motive in lifecycle simulations achieves realistic annuitisation rates. He
solves a simple lifecycle model with bequest motives taken from several recent
papers in the literature. The bequest motives he picks therefore match other
important aspects of the lifecycle model such as how much individuals actually
bequest and how rich individuals are when they bequest.

@@

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Using data from the Association of British
Insurers, they show that annuity demand dropped by 75\% from its maximum in
2011. Figure \ref{fig:annovertime} shows purchases of annuities overtime and
demonstrates the sharp decrease in purchases that happened from 2014 to 2015.
There was an increase in the number of pensions that were being accessed using
an income drawdown product, but these only partly account for the drop in
annuitisation as is shown by the graph.
\begin{figure}[h]
    \caption{New pension pots accessed }
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}
\end{figure}

\begin{figure}[h]
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/annuity_pot_sizes.pdf}
    \label{fig:ann2122}
    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution methods to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most defined contribution pensions were accessed through
annuities.

@@

Table \ref{tab:sum_stats} shows various summary statistics for each variable of
interest from ELSA and the ONS.

@@

Overall, the groups appear to have slightly different wealth profiles with the
treatment group being richer and slightly more likely to have DC pension wealth.
On key demographic characteristics, however, the groups are similar. The
difference in retirement age between the two groups can be mostly explained by
the difference in expected retirement age meaning that individuals have not en
masse decided to delay retirement and avoid annuitisation. Moreover, for the
regression discontinuity models I only use individuals who have a DC pension so
it does not matter that there are more individuals with DC pensions in the
second group.
\begin{landscape}
    \tiny
    \input{tables/r_summary_stats.tex}
    \normalsize
\end{landscape}

@@

\input{tables/cov_balance.tex}

@@

\begin{landscape}
    \input{tables/elsa_results_only_dc.tex}
\end{landscape}

\begin{landscape}
    \input{tables/elsa_results_all_with_interact.tex}
\end{landscape}

\begin{landscape}
    \input{tables/elsa_results_dc_pot_interact.tex}
\end{landscape}

\begin{landscape}
    \input{tables/elsa_results_fin_wealth_interact.tex}
\end{landscape}

@@

Table \ref{tab:DcOnlyFinWealthInteract} contains these results and shows a
similar pattern to Table \ref{tab:DcOnlyInteract}. The reform is associated with
an increase in consumption, but more financial wealth decreases the effect
rather than making it stronger. This could be because those with higher levels
of financial wealth have more assets in other savings apart from retirement
savings \cite{cribb_karjalainen_ifs_2023} and therefore the DC pot is a smaller
proportion of overall assets meaning the policy change did not affect them as
much.

@@

Using the optimal policy function in the last period, which is assets next
period that maximise the utility function and the value function this period
given all the potential income and asset states, I calculate the value of the
last period. This is then used in the problem the year before that. I repeat
this process back to the age of retirement to obtain optimal consumption amounts
for each year of retirement and associated value functions.

To simulate the ELSA data I solve this retirement problem for each new retiree
in the dataset dependent on their objective probability of death each period. I
estimate with subjective life expectancies and objective life expectancies. I
also estimate the model both with and without a bequest motive which was picked
to roughly fit the real annuity rates seen in the data.

@@

Where $C$ is the one-off payment, $\delta$ is the `money's worth' of an annuity
and $p_{t|Retage}$ is the probability of death at age $t$ conditional on being
age $Retage$. So individuals can move C on the asset grid for gaining Ann on the
income grid for the rest of their lives. The `money's worth' of an annuity is
the ratio of expected present discounted value to the price of an annuity that
an individual can purchase. Generally, for annuities, they have been found to be
between $0.8$ and $0.9$ for the population as a whole in the UK,
\cite{finkelstein_porteba_2002} \cite{finkelstein_porteba_2004}.

@@

\begin{landscape}
    \input{tables/standard_life_cycle.tex}
\end{landscape}

\begin{landscape}
    \input{tables/beq_life_cycle.tex}
\end{landscape}

\begin{landscape}
    \input{tables/subjective_life_cycle.tex}
\end{landscape}

@@

@@

          \textbf{Follow up by changing chat because the parameter estimates for
              the diff in diff have changed a bit -- DONE but check again when I
              have gone through it -- haven't checked the comparisons }

    \item \textbf{Abstract}
    \item \textbf{Talk about covariate balance table? We could just set it for
              the DC groups. }
    \item \textbf{Add bit about testing just those people who retired when they
              expected to retire.}
    \item \textbf{Could add consumption and retirement year plot?}
    \item \textbf{Maybe a small discussion of controls for the main regression
              tables as well}
    \item Conclusion
\end{itemize}

>>>>

\hypersetup{ colorlinks, citecolor=black, filecolor=black, linkcolor=black,
    urlcolor=black }
\setlength{\parindent}{0pt}
\usepackage[parfill]{parskip}
\date{July 2023}
\title{Retirement Consumption: Evidence from UK Pension Reform}
\usepackage[letterpaper, margin=1.1in]{geometry}
\linespread{1.5}

\begin{document}
\begin{titlepage}
    \begin{center}

        \normalsize
        {MPhil in Economic Research}
        \vfill

        \huge
        Retirement Consumption: Evidence from UK Pension Reform
        \vfill
        \normalsize
        Candidate number: 5348H

        Date: 27th July, 2023

        Word count: [Insert]
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I find the consumption response of early retirees to the `pensions freedom
    act', which meant defined contribution pensions could be used for other
    options than just purchasing an annuity. I use a discontinuity in year of
    retirement and a difference in difference approach using those who were
    unaffected by the policy reform as a natural control group. I then solve
    three sets of lifecycle models from age of retirement until an agent reaches
    110. I simulate their consumption with and without annuities, i.e. a the
    policy reform, and match their consumption response to what was seen in the
    empirical models. I find that a lifecycle model with a bequest motive best
    fits this consumption response.
\end{abstract}
\newpage
\tableofcontents
\newpage

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \cite{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts - this phenomenon has been termed the "annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature. [\textbf{Add
            something once I have a conclusion}].

@@

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement to
insure against the risk of long life. And more recently
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions.
Since then there has been much literature discussing possible reasons that
people do not annuitise. \cite{finkelstein_porteba_2002,
    finkelstein_porteba_2004} for the UK market and \cite{mitchell_et_al_1999} for
the US market and  find evidence of adverse selection, thereby making the
'money's worth' of annuities lower for the general population as opposed to the
population of annuitants. However, they also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from actuarily fair.

@@

More recently, \cite{lockwood_red_2012} builds on this and shows that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks therefore match other important aspects of the
lifecycle model such as how much individuals actually bequest and how rich
individuals are when they bequest. Also, \cite{odea_sturrock_rest_2023} find that subjective life expectancies can
explain a large part of the annuitisation puzzle especially if agents are less
risk averse.

@@

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Using data from the Association of British
Insurers, they show that annuity demand dropped by 75\% from its maximum in
2011. Figure \ref{fig:annovertime} shows purchases of annuities overtime and
demonstrates the sharp decrease in purchases that happened from 2014 to 2015.
There was an increase in the number of pensions that were being accessed using
an income drawdown product, but these only partly account for the drop in
annuitisation as is shown by the graph.
\begin{figure}[h]
    \caption{New pension pots accessed }
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}
\end{figure}

\begin{figure}[h]
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_pot_sizes.pdf}
    \label{fig:ann2122}
    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution methods to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most defined contribution pensions were accessed through
annuities.

@@

Table \ref{tab:sum_stats}\footnote{There are hyperlinks on the table number so,
    if your pdf viewer supports this, you can click to go to the table or
    figure} shows various summary statistics for each variable of interest from
ELSA and the ONS.

@@

Overall, the groups appear to have slightly different wealth profiles with the
treatment group being richer and slightly more likely to have DC pension wealth.
On key demographic characteristics, however, the groups are similar. The
difference in retirement age between the two groups can be mostly explained by
the difference in expected retirement age meaning that individuals have not en
masse decided to delay retirement and avoid annuitisation. Moreover, for the
regression discontinuity models I only use individuals who have a DC pension so
it does not matter that there are more individuals with DC pensions in the
second group.
\begin{landscape}
    \linespread{1.25}
    \input{tables/r_summary_stats.tex}
    \normalsize
\end{landscape}

@@

\linespread{1.25}
\input{tables/cov_balance.tex}
\linespread{1.5}

@@

\begin{landscape}
    \linespread{1}

    \input{tables/elsa_results_only_dc.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}

    \input{tables/elsa_results_all_with_interact.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}

    \input{tables/elsa_results_dc_pot_interact.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}
    \input{tables/elsa_results_fin_wealth_interact.tex}
\end{landscape}

@@

Table \ref{tab:DcOnlyFinWealthInteract} contains these results and shows a
similar pattern to Table \ref{tab:DcOnlyInteract}. The reform is associated with
an increase in consumption, but more financial wealth decreases the effect
rather than making it stronger. This could be because those with higher levels
of financial wealth have more assets in other savings apart from retirement
savings \citep{cribb_karjalainen_ifs_2023} and therefore the DC pot is a smaller
proportion of overall assets meaning the policy change did not affect them as
much.

\subsection{Robustness Checks}

I carry out two checks for robustness. Firstly, I exclude those individuals who
retired year was greater than a year away from their expected retirement year in
the first interview they gave with ELSA. This checks whether the effects shown
above are being driven by individuals either retiring early to buy cheaper
annuities or retiring later and avoiding annuitisation altogether.

I also check whether the EU reform to annuity pricing is changing the results in
some way. Since the reform only came into affect in December 2012 there are some
retirees in our sample who may have been able to purchase an annuity before
annuity pricing was made the same for men and women. I drop anyone with a
retirement year in 2012 or 2011 and repeat the analysis from Table \ref{tab:DcOnlyRes}.

\begin{landscape}
    \linespread{1}
    \input{tables/rob_elsa_results_only_dc_only_exp.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}
    \input{tables/rob_elsa_results_only_dc_only_late_ret_year.tex}
\end{landscape}

Tables \ref{tab:DcOnlyExpOnlyRes} and \ref{tab:DcOnlyNot2012} show these
results. In Table \ref{tab:DcOnlyExpOnlyRes} we can see the sign is reversed on
PostReform in all consumption categories apart from clothing, however the sample size is a third of what it
was prior to excluding individuals so there is even more uncertainty in these
estimates. Table \ref{tab:DcOnlyNot2012} the sample size is again much smaller
and the sign has flipped on all consumption categories apart from clothing. The
implication of this regression is that those who retired in 2013 had a higher
level of consumption in early retirement than those who retired post 2014.
Although these results give

@@

Using the optimal policy function in the last period, which is assets next
period that maximise the utility function and the value function this period
given all the potential income and asset states, I calculate the value of the
last period. This is then used in the problem the year before that. I repeat
this process back to the age of retirement to obtain optimal consumption amounts
for each year of retirement and associated value functions. I write code in
Julia that solves this maximisation problem.

To simulate the ELSA data I solve this retirement problem for each new retiree
in the dataset dependent on their objective probability of death each period. I
estimate with subjective life expectancies and objective life expectancies. I
also estimate the model both with and without a bequest motive which was picked
to roughly fit the rates of annuitisation that are observed in the real world,
which are around 6\%.

@@

Where $C$ is the one-off payment, $\delta$ is the `money's worth' of an annuity
and $p_{t|Retage}$ is the probability of death at age $t$ conditional on being
age $Retage$. So individuals can move down $C$ on the asset grid for gaining $Ann$ on the
income grid for the rest of their lives. The `money's worth' of an annuity is
the ratio of expected present discounted value to the price of an annuity that
an individual can purchase. Generally they have been found to be
between $0.8$ and $0.9$ for the population as a whole in the UK
\citep{finkelstein_porteba_2002, finkelstein_porteba_2004, mitchell_et_al_1999}.

@@

\begin{landscape}
    \linespread{1.25}

    \input{tables/standard_life_cycle.tex}
\end{landscape}

\begin{landscape}
    \linespread{1.25}

    \input{tables/beq_life_cycle.tex}
\end{landscape}

\begin{landscape}
    \linespread{1.25}

    \input{tables/subjective_life_cycle.tex}
\end{landscape}

@@

I have shown that consumption in early retirement increased overall and
particularly on

@@

          \textbf{Follow up by changing chat because the parameter estimates for
              the diff in diff have changed a bit -- DONE but check again when I
              have gone through it -- haven't checked the comparisons }
    \item \textbf{Abstract}
    \item \textbf{Add bit about testing just those people who retired when they
              expected to retire.}
    \item \textbf{Could add consumption and retirement year plot?}
    \item \textbf{Maybe a small discussion of controls for the main regression
              tables as well}
    \item Conclusion
\end{itemize}

\section{Appendix}
\begin{landscape}
    \input{tables/elsa_results_only_dc_only_exp.tex}
\end{landscape}
